Automating Political Bias Prediction

### 2016-05-24 manuscript/rebuttal.txt

I do hope that the reviewers would consider taking this into account and I hope that it would be possible to add these added references and their discussion to the camera ready version of the manuscript?

An updated version can be found here

I would like to address first the three main concerns of the reviewers and then answer some specific concerns in more detail.

1) Missing related work:
I did extensive literature research, compared and discussed as many studies I could cover and added them to the revised manuscript. My results are well in line with previous findings but extend existing work with new analyses.

2) The analysis only covers german.
That is mostly due to the availability of data - but I will look into more languages asap.

3) Low and badly explained accuracy of predictions:
I compared the results with previous similar studies (Yu et al, "Classifying party affiliation from political speech", 2008 and Hirst et al, "Text to Ideology or Text to Party Status?", 2014).
My results are on par with or better than the results reported in those studies on comparable data. Importantly also the explanations and empirical tests of these accuracies provided in my manuscript were in line with the findings in those studies:

a) the results of binary party classification (see Yu et al, 2008) was comparable (accuracies between 0.8 and 0.87) to the accuracy in the present study (0.85 for the 17th Bundestag)
b) the accuracy in the five class classification of the 17th Bundestag (0.63) is slightly better than the classification accuracy of 0.61 which Hirst et al obtained on a five class problem of political affiliation prediction in the European parliament.
c) As demonstrated in the my submission the main sources of poor accuracy are the out-of-domain generalization challenge and the short length of the texts. Both effects were described before in Yu et al, 2008 and Hirst et al 2014, respectively, in similar (but not the same) settings.

Also it would be worth noting that even if the accuracies are low (as those of comparable studies), they nonetheless are on average above chance. While these results do not allow for usage in production systems for classification, it is well possible to use such a system as assistive technology for human annotators in an active learning setting for priorization of to-be-labeled data.

MORE DETAILED COMMENTS

REVIEWER 2
Q: [...] other classifiers
A: I did try others (random forests, kernel based methods); results on in-domain data were slightly better but results on out of domain data were significantly worse (despite regularization).

@@

Q: What about stemming of words?
A: I did include this in earlier hyperparameter optimizations, but it did not improve results, so i excluded it for brevity.

Q: [sentence level too fine, topic too coarse ]
A: I tried paragraphs for manifestoproject data earlier; results were between sentence and topic performance. Hirst et al, 2014 show a comprehensive analysis of text length and accuracy, though.

@@

Q: [...] 3-fold cross-validation sufficent [...]
A: [...] A single random split into train/hyperparameter-optimization/evaluation is common practice (see e.g. Netflix prize); I did 5 fold nested CV, results were the same.

Q: d is the size of the dictionary [...] union of the words [...] in the corpus and the words [...] in the sentiment dictionary
A: Yes

Q: the sentiment vector s includes all sentiment-related words, both positive and negative.
A: It contains the sentiment polarity values associated with the key words.

Q: What is N in Table 1?
A: Number of data points, the revised manuscript includes this.

### 2016-05-24 manuscript/political_bias_prediction.tex
revision for rebuttal phase

Ideally one would choose for each topic a sample of reports from the entire political spectrum in order to form an unbiased opinion. But ordering media content with respect to the political spectrum at scale requires automated prediction of political bias. The aim of this study is to provide some empirical evidence indicating that leveraging open data sources automated political bias prediction is possible with above chance accuracy.

When human experts determine political bias of texts they will take responsibility for what they say about a text, and they can explain their decisions. This is a key difference to many statistical learning approaches. Not only is the responsibility question problematic, it can also be difficult to interpret some of the decisions. In order to validate the predictions of the models three strategies that allow for better interpretations of the models are proposed. First interpretations of the model misclassifications are related to changes in party policies. Second univariate measures of correlation between text features and party affiliation allow to relate the predictions to the kind of information that political experts use for interpreting texts. Third sentiment analysis is used to investigate whether this aspect of language has discriminatory power.

@@

A similar approach to the one presented here was taken in \cite{Yu2008}. The authors extracted BOW feature vectors and applied linear classifiers to predict political party affiliation of US congress speeches. They used data from the two chambers of the US congress, House and Senat, in order to investigate generalization performance of a classifier trained on one domain and tested on another. They found that accuracies of the model when trained on one domain and test on another was significantly decreased. Generalization was also affected by the time difference between the political speeches used for training and those used for testing.

Other work has focused on developing dedicated methods for predicting political bias such as for instance two methods known as WordFish \cite{Slapin08ascaling} and WordScores \cite{Laver2003}, respectively, and improved versions thereof, see e.g. \cite{Lowe09scalingpolicy}. Validation in these studies is often done by comparing parameter estimates of the different models amongst each other or with human expert annotations. These models are often not evaluated on held-out data or with respect to the same type of labels used to train or partition the data, which renders comparisons with this work or other work such as \cite{Yu2008} difficult. These approaches have been very valuable for a posteriori analysis of historical data but do not seem to be used as much for analyses of new data in a predictive analytics setting. 
%
Also texts from social network data were investigated in the context of automated political text analysis in the 2008 U.S. presidential election \cite{Gryc2014}. The study highlighted the importance of features such as sentiment or social network structure but the results also show that standard BOW features combined with linear classifiers yield competitive results for sentiment classification on political tweets.

Finally Hirst et al conducted a large number of experiments in the Canadian parliament and the European parliament that offer an interesting comparison to the present study both in terms of methodology but also with respect to their results \cite{Hirst2014}. The authors show that a classifier trained on parliament speeches uses language elements of defense and attack to classify speeches, rather than ideological vocabulary. They also argue that emotional content plays an important role in automatic analysis of political texts. Furthermore their results show a clear depencency between length of a political text and the accuracy with which it can be classified correctly.

Taken together, there is a large body of literature in this expanding field in which scientists from quantitative empirical disciplines as well as political science experts collaborate on the challenging topic of automated analysis of political texts. However except for few exceptions most previous work has focused to a large extent on binary classification\footnote{Many parliaments only have two parties and many studies chose binary classification schemes within manually defined topics and generic schemes such as conservative vs liberal or left vs right.} or to assignment of a one dimensional policy position (mostly left vs right). Often this is a too coarse approach to take into account subtle differences in political policies. This work focuses on more fine grained political view prediction: for one the case of the german parliament is more diverse than two parliament systems, allowing for a distinction between more policies; second the political view labels considered are more fine grained than in previous studies. Another important contribution of this work is that many existing studies are primarily concerned with a posteriori analysis of historical data. This work aims at prediction of political bias on out-of-domain data with a focus on the practical application of the model on new data, for which a prototypical web application is provided. The experiments estimate out-of-domain generalization and thus complement the work of \cite{Yu2008, Hirst2014} with results from data of the german parliament. In particular the results of this study confirm and extend some of their findings on why and when generalization performance is impacted negatively.

@@

\section{Conclusions, Limitations and Outlook}\label{sec:conclusion}
This study presents a simple approach for automated political bias prediction. The results of these experiments show that automated political bias prediction is possible with above chance accuracy in some cases. It is worth noting that even if the accuracies are (as those of comparable studies \cite{Yu2008, Hirst2014}) are not perfect, they nonetheless are on average above chance. While these results do not allow for usage in production systems for classification, it is well possible to use such a system as assistive technology for human annotators in an active learning setting.

One of the main limiting factors of an automated political bias prediction system is the availability of training data. Most training data sets that are publicly available have an inherent bias as they are sampled from a different domain. This study tried to quantify the impact of this, but not in all settings out of domain data was available to quantify generalisation performance across text domains. 
For the cases where evaluation data from two domains was available there was a pronounced drop in performance between in domain evaluation set and out of domain evaluation set. This effect was reported previously for similar data, see e.g. \cite{Yu2008}. Also the finding that shorter texts are more difficult to classify than longer texts is in line with previous studies \cite{Hirst2014}. However when considering texts of sufficient length (for instance by aggregating all texts of a given political topic, see \autoref{tab:results_topic}) the classification performance became better; in some cases reliable predictions can be obtained even beyond the training text domain.

Some aspects of these analyses could be interesting for social science researchers; three of these are highlighted here.
First the misclassifications of a model can be related to the changes in policy of a party. Such analyses could be helpful to quantitatively investigate a change in policy. Second analysing the word-party correlations shows that some discriminative words can be related to the political views of a party; this allows for validation of the models by human experts. Third when correlating the sentiment of a speech with proxies of political power there is a strong positive correlation between political power and positive sentiment. While such an insight in itself might seem not very surprising this quantifiable link between power and sentiment could be useful nonetheless: Sentiment analysis is a rather domain independent measure, it can be easily automated and scaled up to massive amounts of text data. Combining sentiment features with other measures of political bias could potentially help to alleviate some of the domain-adaptation problems encountered when applying models trained on parliament data to data from other domains. \\

>>>>

Ideally one would choose for each topic a sample of reports from the entire political spectrum in order to form an unbiased opinion. But ordering media content with respect to the political spectrum at scale requires automated prediction of political bias. The aim of this study is to provide empirical evidence  indicating that leveraging open data sources of german texts, automated political bias prediction is possible with above chance accuracy. These experimental results  confirm and extend previous findings \cite{Yu2008, Hirst2014}; a novel contribution of this work is a proof of concept which applies this technology to sort news article recommendations according to their political bias.

When human experts determine political bias of texts they will take responsibility for what they say about a text, and they can explain their decisions. This is a key difference to many statistical learning approaches. Not only is the responsibility question problematic, it can also be difficult to interpret some of the decisions. In order to validate and explain the predictions of the models three strategies that allow for better interpretations of the models are proposed. First the model misclassifications are related to changes in party policies. Second univariate measures of correlation between text features and party affiliation allow to relate the predictions to the kind of information that political experts use for interpreting texts. Third sentiment analysis is used to investigate whether this aspect of language has discriminatory power.

@@

A similar approach to the one presented here was taken in \cite{Yu2008}. The authors extracted BOW feature vectors and applied linear classifiers to predict political party affiliation of US congress speeches. They used data from the two chambers of the US congress, House and Senat, in order to assess generalization performance of a classifier trained on data from one chamber and tested on data from another. They found that accuracies of the model when trained on one domain and tested on another were significantly decreased. Generalization was also affected by the time difference between the political speeches used for training and those used for testing.

Other work has focused on developing dedicated methods for predicting political bias. Two popular methods are WordFish \cite{Slapin08ascaling} and WordScores \cite{Laver2003}, or improved versions thereof, see e.g. \cite{Lowe09scalingpolicy}. These approaches have been very valuable for {\em a posteriori} analysis of historical data but they do not seem to be used as much for analyses of new data in a predictive analytics setting. Moreover direct comparisons of the results obtained with these so called {\em scaling} methods with the results of the present study or those of studies as \cite{Yu2008} are difficult, due to the different modeling and evaluation approaches: Validations of WordFish/WordScore based analyses often compare parameter estimates of the different models rather than predictions of these models on held-out data with respect to the same type of labels used to train the models.

Finally Hirst et al conducted a large number of experiments on data from the Canadian parliament and the European parliament; these experiments can be directly compared to the present study both in terms of methodology but also with respect to their results \cite{Hirst2014}. The authors show that a linear classifier trained on parliament speeches uses language elements of defense and attack to classify speeches, rather than ideological vocabulary. The authors also argue that emotional content plays an important role in automatic analysis of political texts. Furthermore their results show a clear dependency between length of a political text and the accuracy with which it can be classified correctly.

Taken together, there is a large body of literature in this expanding field in which scientists from quantitative empirical disciplines as well as political science experts collaborate on the challenging topic of automated analysis of political texts. Except for few exceptions most previous work has focused on binary classification\footnote{Many parliaments only have two parties and many studies chose binary classification schemes within manually defined topics and generic schemes such as conservative vs liberal or left vs right.} or on assignment of a one dimensional policy position (mostly left vs right). Yet many applications require to take into account more subtle differences in political policies. This work focuses on more fine grained political view prediction: for one, the case of the german parliament is more diverse than two parliament systems, allowing for a distinction between more policies; second the political view labels considered are more fine grained than in previous studies. While previous studies used such labels only for partitioning training data \cite{Slapin08ascaling} (which is not possible at test time in real-world applications where these labels are not known) the experiments presented in this study directly predict these labels. Another important contribution of this work is that many existing studies are primarily concerned with {\em a posteriori} analysis of historical data. This work aims at prediction of political bias on out-of-domain data with a focus on the practical application of the model on new data, for which a prototypical web application is provided. The experiments on out-of-domain generalization complement the work of \cite{Yu2008, Hirst2014} with results from data of the german parliament and novel sentiment analyses.

@@

\section{Conclusions, Limitations and Outlook}\label{sec:conclusion}
This study presents a simple approach for automated political bias prediction. The results of these experiments show that automated political bias prediction is possible with above chance accuracy in some cases. It is worth noting that even if the accuracies are not perfect, they are above chance and comparable with results of comparable studies \cite{Yu2008, Hirst2014}. While these results do not allow for usage in production systems for classification, it is well possible to use such a system as assistive technology for human annotators in an active learning setting.

One of the main limiting factors of an automated political bias prediction system is the availability of training data. Most training data sets that are publicly available have an inherent bias as they are sampled from a different domain. This study tried to quantify the impact of this effect. 
For the cases in which evaluation data from two domains was available there was a pronounced drop in prediction accuracy between the in domain evaluation set and the out of domain evaluation set. This effect was reported previously for similar data, see e.g. \cite{Yu2008}. Also the finding that shorter texts are more difficult to classify than longer texts is in line with previous studies \cite{Hirst2014}. When considering texts of sufficient length (for instance by aggregating all texts of a given political topic) classification performance improved and in some cases reliable predictions could be obtained even beyond the training text domain.

Some aspects of these analyses could be interesting for social science researchers; three of these are highlighted here.
First the misclassifications of a model can be related to the changes in policy of a party. Such analyses could be helpful to quantitatively investigate a change in policy. Second analysing the word-party correlations shows that some discriminative words can be related to the political views of a party; this allows for validation of the models by human experts. Third when correlating the sentiment of a speech with measures of political power there is a strong positive correlation between political power and positive sentiment. While such an insight in itself might seem not very surprising this quantifiable link between power and sentiment could be useful nonetheless: Sentiment analysis is a rather domain independent measure, it can be easily automated and scaled up to massive amounts of text data. Combining sentiment features with other measures of political bias could potentially help to alleviate some of the domain-adaptation problems encountered when applying models trained on parliament data to data from other domains. \\
